---
title: "Survival Regression Analysis"
author: "Lynle Hayes; Refactored by Andre Gala-Garza"
format: html
editor: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggpubr)
library(survival)
library(survminer)
library(splines)
library(viridis)
library(dplyr)
```

# Learning Objectives
* Visualize cumulative incidence of cancer deaths and other deaths in cancer patients.
* Use Cox proportional hazards models to compare mortality rates among different demographic groups.
* Calculate, visualize and interpret time-varying hazard ratios.

# 1. Import and manage data
```{r}
data <- read.csv("melanoma_survival.csv", header=TRUE)

# Age
table(data$Age.recode.with..1.year.olds.and.90.)

data$age <- dplyr::case_when(
  data[[1]] %in% c("15-19 years","20-24 years","25-29 years","30-34 years",
                   "35-39 years","40-44 years","45-49 years") ~ "<50",
  data[[1]] %in% c("50-54 years","55-59 years") ~ "50-59",
  data[[1]] %in% c("60-64 years","65-69 years") ~ "60-69",
  data[[1]] %in% c("70-74 years","75-79 years","80-84 years","85-89 years","90+ years") ~ "70+",
  TRUE ~ NA_character_
)
data$age <- factor(data$age, levels = c("70+","<50","50-59","60-69"))

# Race
colnames(data)[2] <- "race"
data$race <- factor(data$race)
data$race <- relevel(data$race, ref = "Non-Hispanic White")

# Sex
colnames(data)[3] <- "sex"
data$sex <- factor(data$sex, levels = c("Female","Male"))

# Create Head_and_Neck classification and update subsite factor levels
# Rename column 5 to "subsite"
colnames(data)[5] = "subsite"

# Recode subsite into broader categories
data <- data %>%
  mutate(
    subsite = case_when(
      subsite %in% c("C44.1-Eyelid", "C44.2-External ear",
                     "C44.3-Skin other/unspec parts of face",
                     "C44.4-Skin of scalp and neck") ~ "Head and Neck",
      subsite %in% c("C44.5-Trunk") ~ "Trunk",
      subsite %in% c("C44.6-Upper limb") ~ "Arm",
      subsite %in% c("C44.7-Lower limb") ~ "Leg",
      TRUE ~ subsite
    ),
    subsite = factor(subsite, levels = c("Head and Neck", "Trunk", "Arm", "Leg"))
)

# Year of diagnosis
data$yod <- ifelse(data[[4]] < 2010, "2004-2009", "2010-2022")
data$yod <- factor(data$yod, levels = c("2004-2009","2010-2022"))

# Stage
colnames(data)[6]="stage"
table(data$stage)

data$stage <- dplyr::case_when(
  data$stage %in% c("Blank(s)", "In situ", "Unknown/unstaged/unspecified/DCO") ~ "Unstaged",
  data$stage %in% c("Regional by both direct extension and lymph node involvement",
                    "Regional by direct extension only",
                    "Regional lymph nodes involved only") ~ "Regional",
  data$stage == "Distant site(s)/node(s) involved" ~ "Distant",
  data$stage == "Localized only" ~ "Localized",
  TRUE ~ data$stage
)
data$stage <- factor(data$stage, levels = c("Localized","Regional","Distant","Unstaged"))

# Survival time
colnames(data)[9] <- "time"
data$time <- as.numeric(data$time)
data$time <- data$time / 12  # convert months to years

# Survival event
data$event <- dplyr::case_when(
  data[[7]] == "Dead (attributable to this cancer dx)" ~ "Cancer death",
  data[[8]] == "Dead (attributable to causes other than this cancer dx)" ~ "Other death",
  data[[7]] == "Alive or dead of other cause" & data[[8]] == "Alive or dead due to cancer" ~ "Alive",
  TRUE ~ NA_character_
)
```

# 2. Plot cumulative incidence using the survival package
```{r}
# Create a multistate (competing risks) survival object with no covariates (~1)
sf_HN_MS <- survfit(Surv(time=data$time,event=data$event,type="mstate")~1)
DATA <- matrix(NA,nrow=3*length(sf_HN_MS$time),ncol=3)

# Time for each of "Cancer death", "other death", and "alive"
DATA[,1] <- rep(sf_HN_MS$time,3)

# Probabilities are cumulative. These are where "Cancer death", "other death", and "alive" each start
DATA[,2] <- c(rep(0,length(sf_HN_MS$time)),sf_HN_MS$pstate[,2],sf_HN_MS$pstate[,2]+sf_HN_MS$pstate[,3])

# Probabilities are cumulative. These are where "Cancer death", "other death", and "alive" each end
DATA[,3] <- c(sf_HN_MS$pstate[,2],sf_HN_MS$pstate[,2]+sf_HN_MS$pstate[,3],rep(1,length(sf_HN_MS$time)))
DATA <- as.data.frame(DATA)

colnames(DATA) <- c("x","ymin","ymax")
DATA$state <- c(rep("Cancer Death",length(sf_HN_MS$time)),rep("Other death",length(sf_HN_MS$time)),
             rep("Alive",length(sf_HN_MS$time)))
DATA$state <- factor(DATA$state,levels=c("Cancer Death","Other death","Alive"))

p0 <- ggplot(data = DATA, aes(x = x, y = ymin, ymin = ymin, ymax = ymax, fill = state)) +
  scale_fill_manual(values=c("gray40", "gray2", "grey60"), guide_legend(title = "Age at diagnosis")) +
  geom_ribbon(aes(fill = state,linetype = NA), alpha = 1, show.legend = FALSE) +
  theme_bw(base_size = 8) + labs(x = 'Years since diagnosis', y = 'Fraction') +
  theme(axis.title.y = element_text(vjust=1.5))+
  theme(legend.position.inside = c(0.3,0.8),legend.key = element_blank(),
        legend.background = element_blank(), legend.text = element_text(size=7),
        legend.key.height = unit(0.05,"inch")) +
  theme(panel.border = element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line.x = element_line(color = "black", linewidth = 0.4),
        axis.line.y = element_line(color = "black", linewidth = 0.4))+
  annotate("text", x = c(5, 9, 11), y = c(0.75, 0.05, 0.2), label = c("Alive", "Died from cancer",
                                                                      "Died from other causes"),
                                                            color = "White", size = 2.5) +
  scale_x_continuous(limits = c(0, 15), expand = (c(0.01, 0.01)), breaks = c(0, 5, 10, 15, 20)) + 
  scale_y_continuous(limits = c(0, 1), expand = c(0.005,0.005))

p0
```

## 3. Plotting cumulative incidence by covariate

Here, we create a `survift` object with cumulative incidence stratified by the levels of our variables of interest.
In order to extract information for plotting from the `survfit` objects, we need to know the order of the levels and how many rows are associated with each level.

Let's start by looking at sex and race & ethnicity.

```{r}
data$time <- as.numeric(as.character(data$time))

# Cumulative incidence of cancer death by sex
sf_sex <- survfit(Surv(time=data$time,event=data$event,type="mstate")~data$sex)

# Cumulative incidence of cancer death by race and ethnicity
sf_race_eth <- survfit(Surv(time=data$time,event=data$event,type="mstate")~data$race)

# Cumulative incidence of cancer death by subsite
sf_subsite <- survfit(Surv(time=data$time,event=data$event,type="mstate")~data$subsite)

# Cumulative incidence of cancer death by age
sf_age <- survfit(Surv(time=data$time,event=data$event,type="mstate")~data$age)

# Check the strata of each model
sf_sex$strata
sf_race_eth$strata
sf_subsite$strata
sf_age$strata
```

In the example data, most groups have information across all possible survival months, but that isn't true for small subgroups (in the example data, Non-Hispanic American Indian/Alaskan Native).

```{r}
# Plot cumulative incidence of cancer death by sex
# For ease of plotting for other covariates, use generic plot code after setting
df_sex <- sf_sex
mycolors <- magma

# Create a data matrix with the correct number of rows:
# A 0 for each group plus the sum of the number of rows for each subgroup
DATA_ci_sex <- matrix(NA,nrow=(length(df_sex$strata)+sum(df_sex$strata)),ncol=4)
# x
DATA_ci_sex[,1] <- c(0,df_sex$time[1:df_sex$strata[1]],
           0,df_sex$time[(1+df_sex$strata[1]):(sum(df_sex$strata[1:2]))])
# y
DATA_ci_sex[,2] <- c(0,df_sex$pstate[1:df_sex$strata[1],2],
           0,df_sex$pstate[(1+df_sex$strata[1]):(sum(df_sex$strata[1:2])),2])
# 95% CI (lower bound)
DATA_ci_sex[,3] <- c(0,df_sex$lower[1:df_sex$strata[1],2],
           0,df_sex$lower[(1+df_sex$strata[1]):(sum(df_sex$strata[1:2])),2])
# 95% CI (upper bound)
DATA_ci_sex[,4] <- c(0,df_sex$upper[1:df_sex$strata[1],2],
           0,df_sex$upper[(1+df_sex$strata[1]):(sum(df_sex$strata[1:2])),2])
DATA_ci_sex <- as.data.frame(DATA_ci_sex)
colnames(DATA_ci_sex) <- c("x","y","ymin","ymax")
breaks <- which(c(DATA_ci_sex$x,0)==0)[-1]
DATA_ci_sex$sex <- c(rep("Female",df_sex$strata[1]+1),rep("Male",df_sex$strata[2]+1))
DATA_ci_sex$sex <- factor(DATA_ci_sex$sex,levels=levels(data$sex))

# In this plot, we have two groups, so it gets busy to also show other
# cause deaths. We drop them from this plot.
plot_sex <- ggplot(data=DATA_ci_sex,aes(x = x, y = y, ymin = ymin, ymax = ymax, color = sex)) +
  scale_color_manual(values=mycolors(3)[-3],name ="Sex") +
  geom_ribbon(aes(fill = sex, linetype = NA), alpha = 0.25, show.legend=FALSE) +
  geom_line(linewidth = 0.75) +
  scale_fill_manual(guide = "none", values = mycolors(3)[-3]) +
  theme_bw(base_size = 8) + labs(x = 'Years since diagnosis', y = 'Fraction dead from cancer') +
  theme(axis.title.y = element_text(vjust=1.5)) +
  theme(legend.position = "inside",
        legend.position.inside = c(0.2,0.7), legend.key = element_blank(),
        legend.background = element_blank(), legend.text = element_text(size = 7),
        legend.key.height = unit(0.05,"inch")) +
  theme(panel.border = element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line.x = element_line(color = "black", linewidth = 0.4),
        axis.line.y = element_line(color = "black", linewidth = 0.4)) +
  scale_x_continuous(limits = c(0, 5), expand = (c(0.01, 0.01))) + 
  scale_y_continuous(limits=c(0, 0.1), expand =c(0.005, 0.005))

plot_sex
```

```{r}
# Plot cumulative incidence of cancer death by race and ethnicity
# For ease of plotting for other covariates, use generic plot code after setting
df_re <- sf_race_eth
# Create a data matrix with the correct number of rows:
# A 0 for each group plus the sum of the number of rows for each subgroup.
# Drop "unknown" race.
DATA_ci_re=matrix(NA,nrow=(5+sum(df_re$strata[1:5])),ncol=4)
# x
DATA_ci_re[,1]=c(0,df_re$time[1:df_re$strata[1]],
           0,df_re$time[(1+df_re$strata[1]):(sum(df_re$strata[1:2]))],
           0,df_re$time[(1+sum(df_re$strata[1:2])):(sum(df_re$strata[1:3]))],
           0,df_re$time[(1+sum(df_re$strata[1:3])):(sum(df_re$strata[1:4]))],
           0,df_re$time[(1+sum(df_re$strata[1:4])):(sum(df_re$strata[1:5]))])
# y
DATA_ci_re[,2]=c(0,df_re$pstate[1:df_re$strata[1],2],
           0,df_re$pstate[(1+df_re$strata[1]):(sum(df_re$strata[1:2])),2],
           0,df_re$pstate[(1+sum(df_re$strata[1:2])):(sum(df_re$strata[1:3])),2],
           0,df_re$pstate[(1+sum(df_re$strata[1:3])):(sum(df_re$strata[1:4])),2],
           0,df_re$pstate[(1+sum(df_re$strata[1:4])):(sum(df_re$strata[1:5])),2])
# 95% CI (lower bound)
DATA_ci_re[,3]=c(0,df_re$lower[1:df_re$strata[1],2],
           0,df_re$lower[(1+df_re$strata[1]):(sum(df_re$strata[1:2])),2],
           0,df_re$lower[(1+sum(df_re$strata[1:2])):(sum(df_re$strata[1:3])),2],
           0,df_re$lower[(1+sum(df_re$strata[1:3])):(sum(df_re$strata[1:4])),2],
           0,df_re$lower[(1+sum(df_re$strata[1:4])):(sum(df_re$strata[1:5])),2])
# 95% CI (upper bound)
DATA_ci_re[,4]=c(0,df_re$upper[1:df_re$strata[1],2],
           0,df_re$upper[(1+df_re$strata[1]):(sum(df_re$strata[1:2])),2],
           0,df_re$upper[(1+sum(df_re$strata[1:2])):(sum(df_re$strata[1:3])),2],
           0,df_re$upper[(1+sum(df_re$strata[1:3])):(sum(df_re$strata[1:4])),2],
           0,df_re$upper[(1+sum(df_re$strata[1:4])):(sum(df_re$strata[1:5])),2])

DATA_ci_re=as.data.frame(DATA_ci_re)
colnames(DATA_ci_re)=c("x","y","ymin","ymax")
breaks=which(c(DATA_ci_re$x,0)==0)[-1]
DATA_ci_re$race_eth=c(rep("Non-Hispanic White",   1+df_re$strata[1]),
           rep("Hispanic (All Races)", 1+ df_re$strata[2]),
           rep("Non-Hispanic American Indian/Alaska Native)", 1+ df_re$strata[3]),
           rep("Non-Hispanic Asian or Pacific Islander", 1+ df_re$strata[4]),
           rep("Non-Hispanic Black", 1+ df_re$strata[5]))
DATA_ci_re$race_eth=factor(DATA_ci_re$race_eth)

# In this plot, we have two groups, so it gets busy to also show other
# cause deaths. We drop them from this plot.
plot_race_eth = ggplot(data=DATA_ci_re,aes(x=x,y=y,ymin=ymin,ymax=ymax,color=race_eth))+
  scale_color_manual(values=mycolors(6)[-6],name ="Race and ethnicity")+
  geom_ribbon(aes(fill=race_eth,linetype=NA),alpha=0.25,show.legend=FALSE)+
  geom_line(linewidth=0.75)+
  scale_fill_manual(guide="none",values=mycolors(6)[-6])+
  theme_bw(base_size = 8) + labs(x='Years since diagnosis', y='Fraction dead from cancer')+
  theme(axis.title.y=element_text(vjust=1.5))+
  theme(legend.position = "inside",
        legend.position.inside = c(0.2,0.7),legend.key = element_blank(),
        legend.background=element_blank(),legend.text=element_text(size=7),
        legend.key.height=unit(0.05,"inch"))+
  theme(panel.border = element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line.x = element_line(color = "black",linewidth=0.4),
        axis.line.y = element_line(color = "black",linewidth=0.4))+
  scale_x_continuous(limits=c(0,5),expand=(c(0.01,0.01)))+ 
  scale_y_continuous(limits=c(0,.2),expand=c(0.005,0.005))

plot_race_eth
```

```{r}
# Plot cumulative incidence of cancer death by subsite
# For ease of plotting for other covariates, use generic plot code after setting
df_subsite <- sf_subsite
# Create a data matrix with the correct number of rows:
# A 0 for each group plus the sum of the number of rows for each subgroup.
# Drop "unknown" race.
DATA_ci_subsite=matrix(NA,nrow=(4+sum(df_subsite$strata[1:4])),ncol=4)
# x
DATA_ci_subsite[,1] <- c(0,df_subsite$time[1:df_subsite$strata[1]],
           0,df_subsite$time[(1+df_subsite$strata[1]):(sum(df_subsite$strata[1:2]))],
           0,df_subsite$time[(1+sum(df_subsite$strata[1:2])):(sum(df_subsite$strata[1:3]))],
           0,df_subsite$time[(1+sum(df_subsite$strata[1:3])):(sum(df_subsite$strata[1:4]))])
# y
DATA_ci_subsite[,2] <- c(0,df_subsite$pstate[1:df_subsite$strata[1],2],
           0,df_subsite$pstate[(1+df_subsite$strata[1]):(sum(df_subsite$strata[1:2])),2],
           0,df_subsite$pstate[(1+sum(df_subsite$strata[1:2])):(sum(df_subsite$strata[1:3])),2],
           0,df_subsite$pstate[(1+sum(df_subsite$strata[1:3])):(sum(df_subsite$strata[1:4])),2])
#95% CI (lower bound)
DATA_ci_subsite[,3] <- c(0,df_subsite$lower[1:df_subsite$strata[1],2],
           0,df_subsite$lower[(1+df_subsite$strata[1]):(sum(df_subsite$strata[1:2])),2],
           0,df_subsite$lower[(1+sum(df_subsite$strata[1:2])):(sum(df_subsite$strata[1:3])),2],
           0,df_subsite$lower[(1+sum(df_subsite$strata[1:3])):(sum(df_subsite$strata[1:4])),2])
#95% CI (upper bound)
DATA_ci_subsite[,4] <- c(0,df_subsite$upper[1:df_subsite$strata[1],2],
           0,df_subsite$upper[(1+df_subsite$strata[1]):(sum(df_subsite$strata[1:2])),2],
           0,df_subsite$upper[(1+sum(df_subsite$strata[1:2])):(sum(df_subsite$strata[1:3])),2],
           0,df_subsite$upper[(1+sum(df_subsite$strata[1:3])):(sum(df_subsite$strata[1:4])),2])
DATA_ci_subsite <- as.data.frame(DATA_ci_subsite)
colnames(DATA_ci_subsite) <- c("x","y","ymin","ymax")
breaks <- which(c(DATA_ci_subsite$x,0) == 0)[-1]
DATA_ci_subsite$subsite <- c(rep("Head and Neck", 1+df_subsite$strata[1]),
           rep("Trunk", 1+ df_subsite$strata[2]),
           rep("Arm", 1+ df_subsite$strata[3]),
           rep("Leg", 1+ df_subsite$strata[4]))
DATA_ci_subsite$subsite = factor(DATA_ci_subsite$subsite)

# In this plot, we have two groups, so it gets busy to also show other
# cause deaths. We drop them from this plot.
plot_subsite <- ggplot(data = DATA_ci_subsite, aes(x = x, y = y, ymin = ymin, ymax = ymax, color = subsite)) +
  scale_color_manual(values = mycolors(6)[-6],name ="Anatomical Subsite") +
  geom_ribbon(aes(fill = subsite, linetype = NA), alpha = 0.25, show.legend = FALSE) +
  geom_line(linewidth = 0.75) +
  scale_fill_manual(guide="none",values=mycolors(6)[-6]) +
  theme_bw(base_size = 8) + labs(x='Years since diagnosis', y='Fraction dead from cancer') +
  theme(axis.title.y=element_text(vjust=1.5)) +
  theme(legend.position = "inside",
        legend.position.inside = c(0.2,0.7), legend.key = element_blank(),
        legend.background = element_blank(), legend.text = element_text(size = 7),
        legend.key.height = unit(0.05,"inch")) +
  theme(panel.border = element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line.x = element_line(color = "black", linewidth = 0.4),
        axis.line.y = element_line(color = "black", linewidth = 0.4))+
  scale_x_continuous(limits = c(0,5), expand=(c(0.01, 0.01))) + 
  scale_y_continuous(limits = c(0,0.2), expand=c(0.005, 0.005))

plot_subsite
```

```{r}
# Plot cumulative incidence of cancer death by race and ethnicity
# For ease of plotting for other covariates, use generic plot code after setting
df_age <- sf_age

# Create a data matrix with the correct number of rows:
# A 0 for each group plus the sum of the number of rows for each subgroup.
# Drop "unknown" race.
DATA_ci_age <- matrix(NA,nrow=(4+sum(df_age$strata[1:4])),ncol=4)
# x
DATA_ci_age[,1] <- c(0,df_age$time[1:df_age$strata[1]],
           0,df_age$time[(1+df_age$strata[1]):(sum(df_age$strata[1:2]))],
           0,df_age$time[(1+sum(df_age$strata[1:2])):(sum(df_age$strata[1:3]))],
           0,df_age$time[(1+sum(df_age$strata[1:3])):(sum(df_age$strata[1:4]))])
# y
DATA_ci_age[,2] <- c(0,df_age$pstate[1:df_age$strata[1],2],
           0,df_age$pstate[(1+df_age$strata[1]):(sum(df_age$strata[1:2])),2],
           0,df_age$pstate[(1+sum(df_age$strata[1:2])):(sum(df_age$strata[1:3])),2],
           0,df_age$pstate[(1+sum(df_age$strata[1:3])):(sum(df_age$strata[1:4])),2])
# 95% CI (lower bound)
DATA_ci_age[,3] <- c(0,df_age$lower[1:df_age$strata[1],2],
           0,df_age$lower[(1+df_age$strata[1]):(sum(df_age$strata[1:2])),2],
           0,df_age$lower[(1+sum(df_age$strata[1:2])):(sum(df_age$strata[1:3])),2],
           0,df_age$lower[(1+sum(df_age$strata[1:3])):(sum(df_age$strata[1:4])),2])
# 95% CI (upper bound)
DATA_ci_age[,4] <- c(0,df_age$upper[1:df_age$strata[1],2],
           0,df_age$upper[(1+df_age$strata[1]):(sum(df_age$strata[1:2])),2],
           0,df_age$upper[(1+sum(df_age$strata[1:2])):(sum(df_age$strata[1:3])),2],
           0,df_age$upper[(1+sum(df_age$strata[1:3])):(sum(df_age$strata[1:4])),2])
DATA_ci_age <- as.data.frame(DATA_ci_age)
colnames(DATA_ci_age) <- c("x","y","ymin","ymax")
breaks <- which(c(DATA_ci_age$x,0) == 0)[-1]
DATA_ci_age$age <- c(rep("<50", 1+df_age$strata[1]),
           rep("50-59", 1+ df_age$strata[2]),
           rep("60-69", 1+ df_age$strata[3]),
           rep("+70", 1+ df_age$strata[4]))
DATA_ci_age$age <- factor(DATA_ci_age$age)

# In this plot, we have two groups, so it gets busy to also show other
# cause deaths. We drop them from this plot.
plot_age <- ggplot(data = DATA_ci_age, aes(x = x, y = y, ymin = ymin, ymax = ymax, color = age)) +
  scale_color_manual(values = mycolors(6)[-6], name ="Age Groups") +
  geom_ribbon(aes(fill = age, linetype = NA), alpha = 0.25, show.legend = FALSE) +
  geom_line(linewidth = 0.75) +
  scale_fill_manual(guide="none", values = mycolors(6)[-6]) +
  theme_bw(base_size = 8) + labs(x = 'Years since diagnosis', y = 'Fraction dead from cancer') +
  theme(axis.title.y = element_text(vjust = 1.5)) +
  theme(legend.position = "inside",
        legend.position.inside = c(0.2,0.7),legend.key = element_blank(),
        legend.background = element_blank(), legend.text = element_text(size=7),
        legend.key.height=unit(0.05,"inch")) +
  theme(panel.border = element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line.x = element_line(color = "black", linewidth = 0.4),
        axis.line.y = element_line(color = "black", linewidth = 0.4))+
  scale_x_continuous(limits = c(0, 5), expand = (c(0.01, 0.01)))+ 
  scale_y_continuous(limits = c(0, 0.2), expand = c(0.005, 0.005))

plot_age
```

# 4/ Cox proportional hazards models
We saw in the above plots that covariates impact the survival. If we assume that the rates are proportional over time, what are the hazard ratios?

We will focus on survival in the first 5 years.

```{r}
# Define a censoring variable at 5 years survival
data$censored.time <- ifelse(data$time > 5, 5, data$time)
data$censored.event <- ifelse(data$time > 5, "Alive", as.character(data$event))

# Because we are considering the instantaneous rate of cancer death,
# we treat other deaths as being censored
data$cancer.death <- (data$censored.event == "Cancer death")

# Generate survival object 
cancer.death.surv.5yr <- Surv(data$censored.time, data$cancer.death)

# By sex male
fit.coxph.sex <- coxph(cancer.death.surv.5yr ~ sex,
                       data = data)
summary(fit.coxph.sex)
# The hazard ratio is exp(ceof). Is it significant?

# Stage is the most important predictor of stage, so other variables are 
# vulnerable to confounding. We should adjust for stage.
fit.coxph.sex_stage <- coxph(cancer.death.surv.5yr ~ sex + stage,
                             data = data)
summary(fit.coxph.sex_stage)

fit.coxph.sex_full <- coxph(cancer.death.surv.5yr ~ sex + stage + age + yod + race + event + subsite,
                            data = data)
summary(fit.coxph.sex_full)
# Did the hazard ratio for sex increase or decrease?
# What if we included all the covariates in a multivariable model?
```

```{r}
data_sex_female = data
data_sex_female$sex = relevel(data$sex, ref = "Male")

# By sex female
fit.coxph.sex_female <- coxph(cancer.death.surv.5yr ~ sex,
                              data = data_sex_female)
summary(fit.coxph.sex_female)

# Stage is the most important predictor of stage, so other variable are 
# vulnerable to confounding. We should adjust for stage.
fit.coxph.sex_female_stage <- coxph(cancer.death.surv.5yr ~ sex + stage,
                                    data = data_sex_female)
summary(fit.coxph.sex_female_stage)

fit.coxph.sex_female_full <- coxph(cancer.death.surv.5yr ~ sex + stage + age + yod + race + event + subsite,
                                   data = data_sex_female)
summary(fit.coxph.sex_female_full)
```

```{r}
# Define a censoring variable at 5 years survival
# By race (black)
fit.coxph.race_black <- coxph(cancer.death.surv.5yr ~ race,
                              data = data)
summary(fit.coxph.race_black)
# The hazard ratio is exp(ceof). Is is significant?

# Stage is the most important predictor of stage, so other variable are 
# vulnerable to confounding. We should adjust for stage.
fit.coxph.race_black_stage <- coxph(cancer.death.surv.5yr ~ race + stage,
                                    data = data)
summary(fit.coxph.race_black_stage)

# Stage is the most important predictor of stage, so other variable are 
# vulnerable to confounding. We should adjust for stage.
fit.coxph.race_black_full <- coxph(cancer.death.surv.5yr ~ race + stage + age + yod + sex + event + subsite,
                                   data = data)
summary(fit.coxph.race_black_full)

# Did the hazard ratio for sex increase or decrease?
# What if we included all the covariates in a multivariable model?
```

```{r}
data_race_white = data
data_race_white$race = relevel(data$race, ref = "Non-Hispanic Black")

# Define a censoring variable at 5 years survival
# By race white
fit.coxph.race_white <- coxph(cancer.death.surv.5yr ~ race,
                              data = data_race_white)
summary(fit.coxph.race_white)
# The hazard ratio is exp(ceof). Is is significant?

# Stage is the most important predictor of stage, so other variable are 
# vulnerable to confounding. We should adjust for stage.
fit.coxph.race_white_stage <- coxph(cancer.death.surv.5yr ~ race + stage,
                                    data = data_race_white)
summary(fit.coxph.race_white_stage)

# Stage is the most important predictor of stage, so other variable are 
# vulnerable to confounding. We should adjust for stage.
fit.coxph.race_white_full <- coxph(cancer.death.surv.5yr ~ race + stage + age + yod + sex + event + subsite,
                                   data = data_race_white)
summary(fit.coxph.race_white_full)
# Did the hazard ratio for sex increase or decrease?
# What if we included all the covariates in a multivariable model?
```

```{r}
# Define a censoring variable at 5 years survival
# By subsite
fit.coxph.subsite <- coxph(cancer.death.surv.5yr ~ subsite,
                           data = data)
summary(fit.coxph.subsite)

fit.coxph.subsite_stage <- coxph(cancer.death.surv.5yr ~ subsite + stage,
                                 data = data)
summary(fit.coxph.subsite_stage)

fit.coxph.subsite_full <- coxph(cancer.death.surv.5yr ~ subsite + stage + age + yod + sex + event + race,
                                data = data)
summary(fit.coxph.subsite_full)

```

```{r}
# Define a censoring variable at 5 years survival
# By age
fit.coxph.age <- coxph(cancer.death.surv.5yr ~ age,
                       data = data)
summary(fit.coxph.age)

fit.coxph.age_stage <- coxph(cancer.death.surv.5yr ~ age + stage,
                             data = data)
summary(fit.coxph.age_stage)

fit.coxph.age_full <- coxph(cancer.death.surv.5yr ~ age + stage + subsite + yod + sex + event + race,
                            data = data)
summary(fit.coxph.age_full)
```


```{r}
data_age_70 <- data
data_age_70$age <- relevel(data$age, ref = "<50")

# Define a censoring variable at 5 years survival
# By age greater 70
fit.coxph.age_70 <- coxph(cancer.death.surv.5yr ~ age,
                          data = data_age_70)
summary(fit.coxph.age_70)

fit.coxph.age_70_stage <- coxph(cancer.death.surv.5yr ~ age + stage,
                                data = data_age_70)
summary(fit.coxph.age_70_stage)

fit.coxph.age_70_full <- coxph(cancer.death.surv.5yr ~ age + stage + subsite + yod + sex + event + race,
                               data = data_age_70)
summary(fit.coxph.age_70_full)
```

```{r}
data_subsite_hn <- data
data_subsite_hn$subsite <- relevel(data$subsite, ref = "Trunk")

# Define a censoring variable at 5 years survival
# By race white
fit.coxph.subsite_hn <- coxph(cancer.death.surv.5yr ~ subsite,
                              data = data_subsite_hn)
summary(fit.coxph.subsite_hn)

fit.coxph.subsite_hn_stage <- coxph(cancer.death.surv.5yr ~ subsite + stage,
                                    data = data_subsite_hn)
summary(fit.coxph.subsite_hn_stage)

fit.coxph.subsite_hn_full <- coxph(cancer.death.surv.5yr ~ subsite + stage + age + yod + sex + event + race,
                                   data = data_subsite_hn)
summary(fit.coxph.subsite_hn_full)
```


Censoring the data after 5 years, we compute the Cox proportional hazards models for each covariate in univariable, bivariable (with stage), and multivariable (all covariates) models.

| Group           | Univariable                       | Bivariable (with stage)            | Multivariable                           |     |
|---------------|------------------|------------------|--------------------|-----------------|
| Age \<50 years  | 0.31 (95%CI: 0.29, 0.32); p\<0.001| 0.33 (95%CI: 0.32, 0.34); p\<0.001 | 0.66 (95%CI: 0.63, 0.68); p\<0.001      |     |         
| Age 50-59 years | 0.42 (95%CI: 0.40, 0.43); p\<0.001| 0.44 (95%CI: 0.42, 0.45); p\<0.001 | 0.67 (95%CI: 0.65, 0.69); p\<0.001      |     |          
| Age 60-69 years | 0.52 (95%CI: 0.50, 0.53); p\<0.001| 0.54 (95%CI: 0.52, 0.56); p\<0.001 | 0.72 (95%CI: 0.70, 0.74); p\<0.001      |     | 
| Age 70+ years   | 3.27 (95%CI: 3.15, 3.40); p\<0.001| 3.04 (95%CI: 2.92, 3.16); p\<0.001 | 1.53 (95%CI: 1.45, 1.59); p\<0.001      |     |
| Men             | 1.65 (95%CI: 1.60, 1.69); p\<0.001| 1.49 (95%CI: 1.46, 1.53); p\<0.001 | 0.98 (95%CI: 0.96, 1.01); p = 0.173743  |     |
| Women           | 0.61 (95%CI: 0.59, 0.62); p\<0.001| 0.67 (95%CI: 0.65, 0.69); p\<0.001 | 1.49 (95%CI: 1.46, 1.53); p\<0.001      |     |
| Black           | 2.07 (95%CI: 1.92, 2.23); p\<0.001| 1.34 (95%CI: 1.24, 1.45); p\<0.001 | 1.18 (95%CI: 1.09, 1.27); p\<0.001      |     |          
| White           | 0.48 (95%CI: 0.45, 0.53); p\<0.001| 0.75 (95%CI: 0.69, 0.80); p\<0.001 | 0.85 (95%CI: 0.79, 0.92); p\<0.001      |     |          
| Head and Neck   | 1.62 (95%CI: 1.57, 1.67); p\<0.001| 1.34 (95%CI: 1.30, 1.39); p\<0.001 | 1.00 (95%CI: 0.98, 1.04); p = 0.730132  |     |      
| Trunk           | 0.62 (95%CI: 0.60, 0.64); p\<0.001| 0.74 (95%CI: 0.72, 0.77); p\<0.001 | 0.99 (95%CI: 0.96, 1.03); p = 0.730132  |     |     
| Arm             | 0.52 (95%CI: 0.50, 0.54); p\<0.001| 0.66 (95%CI: 0.64, 0.68); p\<0.001 | 0.92 (95%CI: 0.89, 0.95); p\<0.001      |     |  
| Leg             | 0.74 (95%CI: 0.72, 0.77); p\<0.001| 0.79 (95%CI: 0.76, 0.81); p\<0.001 | 0.92 (95%CI: 0.79, 0.95); p\<0.001      |     |     
